---
title: "Metabolomics Limma"
author: "*Vickie*"
date: "*`r format(Sys.time(), '%d %B, %Y')`*"
output: 
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    theme: spacelab
editor_options: 
  chunk_output_type: console
---
```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, include=FALSE, fig.align = 'center', fig.width = 10, fig.height = 7)
```


```{r}
#Load things
library(sets)
library(useful)
library(taigr)
library(magrittr)
library(plyr)
library(tidyverse)
library(glmnet)
library(ranger)
library(pheatmap)
library(RColorBrewer)
library(prismr)
library(piano)
library(cdsr)
library(ggthemes)
library(scales)
#library(GSEABase)
library(hrbrthemes)
library(colorspace)
library(ggrepel)
library(cowplot)
library(gridExtra)
library(edgeR)
library(reshape2)
library(WGCNA)
library(plotly)
library(limma)
library(edgeR)
library(biomaRt)
library(reshape)
```



```{r}
#Read in data
#data <- data.table::fread(file = "./Data/X0090_BRD1748_raw_data_composite-2019-02-07.csv")
data <- data.table::fread(file = "./Data/19_0425_RONG_CERS1_CP_Results_V2_1.csv")
```



```{r}
#Check total counts for each condition (column in data)

#dat <- data %>% mutate(rownames = makeUnique(compound)) %>% column_to_rownames(var = "rownames") %>% dplyr::select(-compound)
#dat <- data %>% column_to_rownames(var = "compound") #%>% dplyr::select(-compound)
dat <- data %>% as.data.frame %>% dplyr::select(-starts_with("PREF")) %>% 
    dplyr::mutate(Compound = ifelse(Metabolite == "", Compound, Metabolite)) %>% #column_to_rownames(var = "Compound") %>%
    dplyr::select(-MZ,-RT,-HMDB,-Assignment_certainty,-Metabolite)
#dat[dat < 4000 | is.na(dat)] <- 4000   # setting mincount to 4000
dat %<>% #dplyr::filter(rowSums(is.na(dat))<40) %>%
    column_to_rownames(var = "Compound")
dat[is.na(dat) | dat < 1000] <- 1000
colnames(dat) %<>% str_replace_all('_CERS1', '-CERS1') 

x <- NA
x <- DGEList(counts = dat)
print(dim(x))

#x also stores library size, which is just sum of read counts over rows
stopifnot(cbind(colSums(x$counts) == x$samples$lib.size))

head(x$counts) %>% DT::datatable()

x$samples$group <- x$samples %>% rownames() %>% word(sep = fixed("_"), 1,3)
x$samples$cellline <- x$samples$group %>% word(sep = fixed("_"),1)
x$samples$treatment <- x$samples$group %>% word(sep = fixed("_"),2)
x$samples$timepoint <- x$samples$group %>% word(sep = fixed("_"),3)
x$samples$replicate <- x$samples %>% rownames() %>% word(sep = fixed("_"), 4)

x$samples %>% DT::datatable()

metid <- rownames(x)
x$mets <- metid %>% as.data.frame %>% set_colnames("met")#package gene information along with DGE object
x$mets %>% head %>% DT::datatable()
```

```{r}
#prior_count <- 1
g1 <- log10(x$counts) %>% #logCPM
    melt() %>% #convert to long-form (for input to ggplot)
    ggplot(aes(value, color = X2)) + #make a separate density plot for each sample
        geom_density(show.legend = F) +
        xlab('log CPM')
g1

x$counts[x$counts == 0] <- NA
lcpm <- log2(x$counts)
```

```{r}
g1 <- ggplot(melt(lcpm), aes(2^value)) + xlab('CPM') + geom_histogram() + ggtitle('CPM')
g2 <- ggplot(melt(lcpm), aes(value)) + xlab('log CPM') + geom_histogram() + ggtitle('Log CPM')
plot_grid(g1, g2, ncol=2)

plotMDS(lcpm, labels=x$samples$group)
plotMDS(lcpm, labels=x$samples$cellline)
plotMDS(lcpm, labels=x$samples$treatment)
plotMDS(lcpm, labels=x$samples$timepoint)
plotMDS(lcpm, labels=x$samples$replicate)
```


```{r}
sampleDists <- as.matrix(dist(t(lcpm)))
annotation_col <- data.frame(group = x$samples$group, cellline = x$samples$cellline, treatment = x$samples$treatment, timepoint = x$samples$timepoint, replicate = x$samples$replicate) %>% set_rownames(rownames(x$samples))
pheatmap::pheatmap(sampleDists,
                   annotation_col = annotation_col,
                   color = colorRampPalette(c("navy", "white", "firebrick3"))(50))

#rownames(x$samples) <- make.names(rownames(x$samples))
#colnames(x$counts) <- make.names(colnames(x$counts))

x$samples$groupnames <- make.names(x$samples$group)
design <- model.matrix(~0 + groupnames, x$samples)
colnames(design) <- gsub("groupnames", "", colnames(design))
design

contr.matrix <- makeContrasts(
   `639V_BRD-1748_4Hours::639V_DMSO_4Hours` = `X639V_BRD.1748_4Hours` - `X639V_DMSO_4Hours`, 
   `639V_P059_4Hours::639V_DMSO_4Hours` = `X639V_P059_4Hours` - `X639V_DMSO_4Hours`, 
   `639V_BRD-1748-P059_4Hours::639V_BRD-1748_4Hours` = `X639V_BRD.1748.P059_4Hours` - `X639V_BRD.1748_4Hours`, 
   `639V_BRD-1748-P059_4Hours::639V_P059_4Hours` = `X639V_BRD.1748.P059_4Hours` - `X639V_P059_4Hours`, 
   `CJM_BRD-1748_4Hours::CJM_DMSO_4Hours` = `CJM_BRD.1748_4Hours` - `CJM_DMSO_4Hours`, 
   `CJM_BRD-1748-P059_4Hours::CJM_BRD-1748_4Hours` = `CJM_BRD.1748.P059_4Hours` - `CJM_BRD.1748_4Hours`, 
   `CJM-CERS1_BRD-1748_4Hours::CJM-CERS1_DMSO_4Hours` = `CJM.CERS1_BRD.1748_4Hours` - `CJM.CERS1_DMSO_4Hours`, 
   `CJM-CERS1_BRD-1748-P059_4Hours::CJM-CERS1_BRD-1748_4Hours` = `CJM.CERS1_BRD.1748.P059_4Hours` - `CJM.CERS1_BRD.1748_4Hours`, 
   `CJM_BRD-1748_4Hours::CJM-CERS1_BRD-1748_4Hours` = `CJM_BRD.1748_4Hours` - `CJM.CERS1_BRD.1748_4Hours`, 
   `639V_BRD-1748_8Hours::639V_DMSO_8Hours` = `X639V_BRD.1748_8Hours` - `X639V_DMSO_8Hours`, 
   `639V_P059_8Hours::639V_DMSO_8Hours` = `X639V_P059_8Hours` - `X639V_DMSO_8Hours`, 
   `639V_BRD-1748-P059_8Hours::639V_BRD-1748_8Hours` = `X639V_BRD.1748.P059_8Hours` - `X639V_BRD.1748_8Hours`, 
   `639V_BRD-1748-P059_8Hours::639V_P059_8Hours` = `X639V_BRD.1748.P059_8Hours` - `X639V_P059_8Hours`, 
   `CJM_BRD-1748_8Hours::CJM_DMSO_8Hours` = `CJM_BRD.1748_8Hours` - `CJM_DMSO_8Hours`, 
   `CJM_BRD-1748-P059_8Hours::CJM_BRD-1748_8Hours` = `CJM_BRD.1748.P059_8Hours` - `CJM_BRD.1748_8Hours`, 
   `CJM-CERS1_BRD-1748_8Hours::CJM-CERS1_DMSO_8Hours` = `CJM.CERS1_BRD.1748_8Hours` - `CJM.CERS1_DMSO_8Hours`, 
   `CJM-CERS1_BRD-1748-P059_8Hours::CJM-CERS1_BRD-1748_8Hours` = `CJM.CERS1_BRD.1748.P059_8Hours` - `CJM.CERS1_BRD.1748_8Hours`,
   `CJM_BRD-1748_8Hours::CJM-CERS1_BRD-1748_8Hours` = `CJM_BRD.1748_8Hours` - `CJM.CERS1_BRD.1748_8Hours`, 
   levels = colnames(design))
contr.matrix
```

```{r}
#x$samples$lib.size <- 1e6 - 1
#voom(x, design, plot=TRUE)
vfit <- lmFit(log2(x$counts), design) #fits the model
vfit <- contrasts.fit(vfit, contrasts=contr.matrix) #extracts the info we need for the comparisons we're interested in
efit <- eBayes(vfit, trend = T) #This step allows the model to 'pool' information across genes and gain statistical power
plotSA(efit, main="Final model: Mean−variance trend")

summary(decideTests(efit))
# 
# par(mfrow=c(1,2))
# x$samples$lib.size <- 1
# v <- voom(x, design, plot=TRUE) #this operates on the normalized rounts data
# vfit <- lmFit(v, design) #fits the model
# vfit <- contrasts.fit(vfit, contrasts=contr.matrix) #extracts the info we need for the comparisons we're interested in
# efit <- eBayes(vfit) #This step allows the model to 'pool' information across genes and gain statistical power
# plotSA(efit, main="Final model: Mean−variance trend")
# par(mfrow=c(1,1))
# 
# summary(decideTests(efit))
```

### Differentially expressed metabolites {.tabset}

We use limma to identify differentially expressed metabolites between pairs of conditions of interest. Each tab is labeled A::B, denoting the difference A - B. Each tab contains a volcano plot of the metabolites as well as the table used to generate the plot. The p-values are adjusted for multiple testing using the Benjamini-Hochberg procedure.

```{r, include = T}
out <- ''
contrastslist <- colnames(contr.matrix)
for (contrast in contrastslist) {
    out = c(out, knitr::knit_expand(file = './markdown_helpers/limma_detable.Rmd'))
}
```

`r paste(knitr::knit(text = out), collapse = '\n')`

```{r, eval = F}
#Check things
x$counts %>% 
    melt() %>% 
    dplyr::rename("metabolite" = X1, "condition" = X2) %>%
    dplyr::mutate(condition = word(condition, sep = fixed("_"), 1, 3)) %>% 
    dplyr::filter(metabolite == "QI5721") %>%
    ggplot() + geom_boxplot(aes(x = condition, y = log2(value))) +
    coord_flip()
```

```{r, eval = F}
setwd("~/Documents/CalicoHITTA/Calico-Hit-Advancement/BRD-1748/limma_detables/")
bigtable <- NA
for(fname in list.files("~/Documents/CalicoHITTA/Calico-Hit-Advancement/BRD-1748/limma_detables/")) {
    #fname <- list.files("~/Documents/CalicoHITTA/Calico-Hit-Advancement/BRD-1748/limma_detables/")[1]
    f <- data.table::fread(fname) %>% as.data.frame
    f %<>% dplyr::select(compound, logFC, t, P.Value, adj.P.Val) 
    
    colnames(f)[2:5] <- paste0(word(fname,sep = fixed(".")),".",colnames(f)[2:5])
    if(is.na(bigtable)) {
        bigtable <- f
    } else {
        bigtable %<>% left_join(f)
    }
}

write_csv(bigtable, "~/Documents/CalicoHITTA/Calico-Hit-Advancement/BRD-1748/combined_detables.csv")

metabolites <- rownames(dat)
dattidy <- dat %>% t %>% as.data.frame %>% 
    rownames_to_column(var = "sample") %>% 
    dplyr::mutate(sample = str_replace_all(.$sample, '_CERS1', '-CERS1')) %>%
    dplyr::mutate(group = word(sample,1,3,sep = fixed("_"))) %>%
    # dplyr::mutate(cellline = word(sample,sep = fixed("_")), trt = word(sample,2,sep = fixed("_")),
    #               time = word(sample,3,sep = fixed("_")),
    #               rep = word(sample,4,sep = fixed("_"))) %>%
    dplyr::select(-sample) %>%
    tidyr::gather(metabolites, key = "mtb",value = "count")

datstats <- dattidy %>% dplyr::group_by(group, mtb) %>% 
    dplyr::summarize(mean = mean(count), stddev = sqrt(var(count)))

meantable <- datstats %>% 
    dplyr::select(-stddev) %>% 
    spread(key = group, value = mean) %>% 
    column_to_rownames(var = "mtb") %>%
    set_colnames(paste0(colnames(.),".mean")) %>%
    rownames_to_column(var = "compound")

stddevtable <- datstats %>% 
    dplyr::select(-mean) %>% 
    spread(key = group, value = stddev) %>% 
    column_to_rownames(var = "mtb") %>%
    set_colnames(paste0(colnames(.),".stddev")) %>%
    rownames_to_column(var = "compound")

#bigtable %>% left_join(meantable) %>% left_join(stddevtable) %>% write_csv(path = "~/Documents/CalicoHITTA/Calico-Hit-Advancement/BRD-1748/combined_detables_stats.csv")
```

```{r, eval = F}
cdsr::cds_render(path_to_rmd = "~/Documents/CalicoHITTA/Calico-Hit-Advancement/BRD-1748/metabolomics_limma.Rmd", path_to_html = "~/Documents/CalicoHITTA/Calico-Hit-Advancement/BRD-1748/metabolomics_limma.html", folder = "vwang")
```
